### 2022-06-27 main/chapter_2ljets.tex
expand fnp intro, thank eirik, improve text

\TODO{composition (INT)}
% components
% lots of non-prompt, a little fake hadrons
%

Historical precedent leads us to use the `Matrix Method' to estimate
backgrounds including fake and non-prompt leptons, so named because it works by
estimating a linear relationship between dilepton event rates.
\TODO{previous 2ljets MM}

In data, one measures tightly ($t$) and loosely ($l$) identified leptons, leading
to four categories of events with two ordered leptons as labelled $tt$, $tl$,
$lt$, and $ll$ for tight-tight, tight-loose, and so on.
The Matrix Method assumes a linear relationship between these selections
and those of real ($r$) and fake/non-prompt ($f$) leptons:
\begin{equation}
\begin{pmatrix}
\mu^\textrm{data}_{tt} \\
\mu^\textrm{data}_{tl} \\
\mu^\textrm{data}_{tl} \\
\mu^\textrm{data}_{ll} \\
\end{pmatrix}
=
M
\begin{pmatrix}
\mu^\textrm{truth}_{rr} \\
\mu^\textrm{truth}_{fr} \\
\mu^\textrm{truth}_{rf} \\
\mu^\textrm{truth}_{ff} \\
\end{pmatrix}
,
\end{equation}
where $\mu^{\ldots}_{**}$ is a Poisson event rate and $M$ is the matrix of
transition probabilities after which this method is
named~\cite{ATLAS-CONF-2014-058}.

@@

Transition probabilities $p(*\!\mid\!*,i)$ are estimated by studies in
separate selections and assigned as functions of the lepton
flavour, $\pt$ and $|\eta|$.
Given those estimated transition probabilities and some observed data
in a selection of two loose leptons,
$n^\textrm{data}_{tt}$,
$n^\textrm{data}_{tl}$,
$n^\textrm{data}_{tl}$,
and $n^\textrm{data}_{ll}$,
the Matrix Method inverts $M$ to estimate truth level rates
$\bar \mu^\textrm{truth}_{**}$.
(Recall that tight implies loose, so this selection of loose leptons includes
some tight leptons.)
Finally, these truth-level estimates are converted back to data-level
estimates in the tight-tight selection of our search regions back through the
independent transition probabilities as
\begin{align}
% rr
\bar \mu^\textrm{data}_{tt\mid rr} &=
p(t\!\mid\!r,1)
\times p(t\!\mid\!r,2)
\times \bar \mu^\textrm{truth}_{rr}~~, \\
% fr
\bar \mu^\textrm{data}_{tt\mid fr} &=
p(t\!\mid\!f,1)
\times p(t\!\mid\!r,2)
\times \bar \mu^\textrm{truth}_{fr}~~, \\
% rf
\bar \mu^\textrm{data}_{tt\mid rf} &=
p(t\!\mid\!r,1)
\times p(t\!\mid\!f,2)
\times \bar \mu^\textrm{truth}_{rf}~~,~~\textrm{and} \\
% ff
\bar \mu^\textrm{data}_{tt\mid ff} &=
p(t\!\mid\!f,1)
\times p(t\!\mid\!f,2)
\times \bar \mu^\textrm{truth}_{ff}~~.
\end{align}

To assign fake/non-prompt estimates in our regions, this method is
implemented by assigning weights to observed data events, which are
histogrammed in the tighter selections of our search
regions~\cite{twoljets2018int}.
Systematic variations are assigned by comparing histograms when those weights
are varied according to various uncertainties in the Matrix Method procedure.

Yes, that is a histogramming of real data with weights, and those data include
tight-tight events which comprise the signal data of our search regions,
in which we are searching for anomalies.
Since it includes real data, this method looks at all unblinded data, including
signal regions.
And because it knows the data we are supposed to be predicting, it is no
prediction at all!

That last statement was an exaggeration.
Although it does cheat by using the data we should be blind to, it uses those
data in a diluted way, such that they which might end up not making much
difference.

In fact, real data tend to be assigned negative Matrix Method weights.
For example, CR-VZ receives $157$ weighted data in its Matrix Method sample,
of which $135$ pass the signal lepton requirements (tight-tight)%
\footnote{%
I am not sure why this differs from the $194$ actual data observed in CR-VZ.
The two do overlap in event identity numbers, so there is presumably some
tighter selection in the Matrix Method which should be accounted for in its
weights.
},
and the remainder have one loose lepton (zero loose-loose lepton events
are observed here).
The signal events have an average weight of $-0.0227$, and the others
have an average weight of $+0.238$ (one can check these recover the central
fake/non-prompt background of $2.2$).
Such negative signal weights suggest that a local \emph{excess} of data
would act to \emph{decrease} the Matrix Method estimate in the region.
Yes this backwards dependency is backwards, but this gives some relief that it
should not hide real local excesses.

@@

% TODO cite past searches
% TODO thank Eirik

Figure~\ref{fig:2ljets_vrss}
Table~\ref{tab:2ljets_fnp_negative_mm}
Table~\ref{tab:2ljets_fnp_estimates}

>>>>

Given our requirements on lepton quality, isolation, and hardness, it is
expected that the rates of fake leptons, or non-prompt leptons (which do not
originating from the hard scatter) should be small.
And indeed more details estimates indicate that they are small.
However, just like other backgrounds, it is important that we assign
honest and honestly uncertain estimates for these fake/non-prompt backgrounds.

Although my style of spelling out fake/non-prompt every time
(as opposed to just `fake') may be awkward, it is supported by fact that this
the non-prompt part dominates.
Indeed, figures in our internal documentation~\cite{twoljets2018int} suggest
that, in simulations, non-prompt hadron decays produce in excess of $90\%$ of
this background in a loose selections near to our regions, and across the
entire $\metsig$ spectrum.
Non-prompt muons are mostly from decays of heavy-flavour ($b$ or $c$) hadrons,
and electrons arise in roughly equal parts from light flavour hadrons.
Tau decays form the next largest part, with prompt fakes due to misidentified
photons or leptons in the sub-$\%$ weeds.

However, simulations are not totally accurate.
In particular, do not use simulations of pure hadronic processes, so do not
really know the rates at which these strong processes can fake leptons.
Although prompt strong effects may only rarely fake leptons, they have very
large cross-sections are not modelled well.
This motivates the use of data-driven fake estimates to replace the need for
such specialized simulations of poorly understood physics.

Precedent from previous searches in this channel~\cite{
atlas_2l_SUSY_2013_11,
atlas_23l_SUSY_2016_24
}
leads us to use the `Matrix Method'~\cite{ATLAS-CONF-2014-058} to estimate
backgrounds of fake and non-prompt leptons, so named because it works by
estimating a linear relationship between dilepton event rates, which is
written as a matrix multiplication.%
\footnote{%
To be explicit, the Matrix Method does not use Matrix Elements from quantum
scattering, and is not related to the similarly-named
Matrix Element Method~\cite{gainer2013matrix}.
}

\emph{
I am deeply grateful to Eirik~Gramstad for developing the Matrix Method for our
$\twoljets$ searches, documenting it, providing its samples, and showing us how
to use them.
Although I have reservations about the method itself (expounded below), these
are not criticisms of his excellent work spent in using it.
}

In data, one measures tightly ($t$) and loosely ($l$) identified leptons, leading
to four categories of events with two ordered leptons as labelled $tt$, $tl$,
$lt$, and $ll$ for tight-tight, tight-loose, and so on.
The Matrix Method assumes a linear relationship between these selections
and those of real ($r$) and fake/non-prompt ($f$) leptons:
\begin{equation}
\label{eqn:2ljets_matrix_method}
\begin{pmatrix}
\mu^\textrm{data}_{tt} \\
\mu^\textrm{data}_{tl} \\
\mu^\textrm{data}_{tl} \\
\mu^\textrm{data}_{ll} \\
\end{pmatrix}
=
M
\begin{pmatrix}
\mu^\textrm{truth}_{rr} \\
\mu^\textrm{truth}_{fr} \\
\mu^\textrm{truth}_{rf} \\
\mu^\textrm{truth}_{ff} \\
\end{pmatrix}
,
\end{equation}
where $\mu^{\ldots}_{**}$ is a Poisson event rate and $M$ is the matrix of
transition probabilities after which this method is
named~\cite{ATLAS-CONF-2014-058}.

@@

Transition probabilities $p(*\!\mid\!*,i)$ are estimated by studies in
separate selections and assigned as functions of the lepton
flavour, $\pt$ and $|\eta|$.
Given those estimated transition probabilities, the Matrix Method begins by
taking some observed data
$n^\textrm{data}_{tt}$,
$n^\textrm{data}_{tl}$,
$n^\textrm{data}_{tl}$,
and $n^\textrm{data}_{ll}$
in a selection of two loose leptons.
(Recall that tight implies loose, so this selection of loose leptons includes
some tight leptons.)
The Matrix method uses these data as estimate $\mu^\textrm{data}_{**}$,
then inverts $M$ to estimate the truth level rates
$\bar \mu^\textrm{truth}_{**}$ through Equation~\ref{eqn:2ljets_matrix_method}.

Finally, these truth-level estimates are converted back to data-level
estimates in the tight-tight selection of our search regions back through the
independent transition probabilities as
\begin{align}
% rr
\bar \mu^\textrm{data}_{tt\mid rr} &=
p(t\!\mid\!r,1)
\times p(t\!\mid\!r,2)
\times \bar \mu^\textrm{truth}_{rr}~~, \\
% fr
\bar \mu^\textrm{data}_{tt\mid fr} &=
p(t\!\mid\!f,1)
\times p(t\!\mid\!r,2)
\times \bar \mu^\textrm{truth}_{fr}~~, \\
% rf
\bar \mu^\textrm{data}_{tt\mid rf} &=
p(t\!\mid\!r,1)
\times p(t\!\mid\!f,2)
\times \bar \mu^\textrm{truth}_{rf}~~,~~\textrm{and} \\
% ff
\bar \mu^\textrm{data}_{tt\mid ff} &=
p(t\!\mid\!f,1)
\times p(t\!\mid\!f,2)
\times \bar \mu^\textrm{truth}_{ff}~~.
\end{align}

To assign fake/non-prompt estimates in our regions, the Matrix Method is
implemented by assigning weights to observed data events, and histogramming
them in the tighter selections of our search
regions~\cite{twoljets2018int}.
Systematic variations are assigned by comparing histograms when those weights
are varied according to various uncertainties in the Matrix Method procedure.

Yes, actual data are histogrammed in the Matrix Method, and those data include
tight-tight events which comprise the data of our search regions, and those
are the same data in which we are searching for anomalies.
Since it includes actual data, this method unblinds all regions,
including signal regions, early.
And because it knows the data we are supposed to be predicting, it is no
prediction at all!

That last statement was a slight exaggeration.
Although it does cheat by using the data we should be blind to, it uses those
data in a diluted way, such that they can end up not making much difference.

In fact, real data tend to be assigned negative Matrix Method weights.
For example, CR-VZ receives $157$ data in its Matrix Method sample,%
\footnote{%
This count of $157$ differs from the $194$ actual data observed in CR-VZ.
I do not know why.
The two do overlap in event identity numbers, so there is presumably some
tighter selection in the Matrix Method which should be accounted for in its
weights.
}
of which $135$ pass the signal lepton requirements (tight-tight),
and the remainder have one loose lepton (zero loose-loose lepton events
are observed in CR-VZ).
The tight-tight events have an average weight of $-0.0227$, and the others
have an average weight of $+0.238$ (one can check these recover the central
fake/non-prompt background of $2.2$).
Such negative signal weights suggest that a local \emph{excess} of data
would act to \emph{decrease} the Matrix Method estimate in the region.
Yes this backwards dependency is backward, but at least it gives some relief
hide real local excesses would not be hidden by increased fake estimates.

@@

% TODO
Figure~\ref{fig:2ljets_vrss}
Table~\ref{tab:2ljets_fnp_negative_mm}
Table~\ref{tab:2ljets_fnp_estimates}
